### 2017-12-20 chapters/introduction.tex
Fix some final typos

\section{Aim of thesis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PAST %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cardiovascular diseases are the leading cause of death in the
world, responsible for more than 17.3 million deaths annually, a
number that is expected to reach 23.6 million by 2030
\cite{writing2016heart}. Advancements in detection and treatment
have reduced morbidity and mortality, but knowledge of the underlying
mechanisms for the development of the disease, is still very poor.
Consequently, while some patients respond very well to a therapy, others suffering
from the same symptoms might not respond at all.
Our understanding of these mechanisms has been improved through
development of advance imaging techniques and through
computational modeling. While imaging plays a central role in
assessment of cardiac structure and kinematics, biophysical models can
help us to understand causality in this multiscale environment.

@@

A paradigm shift in the field of cardiac mechanics modeling came in
the late 1980s at the University of Auckland. Here the first models
based on a continuum description of the heart were developed
\cite{hunter1988analysis}, which is still are use today.

>>>>

\section{Aim of thesis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PAST %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cardiovascular diseases are the leading cause of death in the
world, responsible for more than 17.3 million deaths annually, a
number that is expected to reach 23.6 million by 2030
\cite{writing2016heart}. Advancements in detection and treatment
have reduced morbidity and mortality, but knowledge of the underlying
mechanisms for the development of the disease, is still very poor.
Consequently, while some patients respond very well to a therapy, others suffering
from the same symptoms might not respond at all.
Our understanding of these mechanisms has improved through
development of advance imaging techniques and through
computational modeling. While imaging plays a central role in
assessment of cardiac structure and kinematics, biophysical models can
help us to understand causality in this multiscale environment.

@@

A paradigm shift in the field of cardiac mechanics modeling came in
the late 1980s at the University of Auckland. Here the first models
based on a continuum description of the heart were developed
\cite{hunter1988analysis}, which are still used today.

### 2017-12-20 chapters/introduction/mechanics.tex
Fix some final typos

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is, the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T = \mathbf{a}_0\otimes\mathbf{a}_0 \},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
Some of the invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

@@

The myocardium contains small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts, this
perfused blood is squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. A material that change its
volume in response to applied loads are referred to as compressible. When the 
volume is preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac mechanical modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompressible. The reason for this choice is
purely numerical.

@@

\subsection{Force-balance equation}
We will now collect all the terms that are involved in the force balance for
the cardiac mechanics problem. Considering the myocardium as an incompressible,
hyperelastic material we obtain the following strong form in the
Lagrangian formulation 
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derive weak
variational form of this equation.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

In this thesis we have in used a transversely isotropic version of
\eqref{eq:holzapel_full} which is obtained by setting $a_{fs} =
b_{fs}= a_s = b_s = 0$, i.e 
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces (after a
series expansion of the exponential and a limiting argument) to
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    J\Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right),
  \end{split}
\end{align}
where $\mathbf{B} = \F \F^{T}$ is the left Cauchy-Green tensor,
$\mathbf{f} = \F \ef$ and $\mathbf{s} = \F \eS$.

@@

The active part $\F_a$, is an inelastic process driven by the
biochemistry and can be seen as the actual distortion of the
microstructure. The elastic part $\F_e$ is responsible for preserving
compatibility of the tissue and stores all the energy in the
deformations. As a consequence, the strain energy function is a
function of the elastic deformation gradient only. The
decoupling can illustrated by considering two sarcomeres connected in
series as shown in Figure \ref{fig:actstrain}.

@@

The FEniCS project is an open-source computing platform for solving
partial differential equations (PDEs) using the finite element method (FEM).
Solving PDEs using FEM involves many implementation details that can
be tedious to implement yourselves. The idea behind FEniCS is to
automate code generation so that the user can spend more time on doing
research and less time on implementation of assembly matrices. At the
core of FEniCS is DOLFIN \cite{logg2012dolfin}, which is C++/Python
library, and works as the main interface in FEniCS. In this thesis
only the Python interface has been used, in which C++ code is
automatically generated using SWIG. This allows for simplicity through
the Python scripting language and the speed of the C++ language.
The domain specific language used to represent weak formulations is
called the Unified Form Language (UFL) \cite{alnaes2014unified}, and
allows for e.g automatic differentiation of forms and expressions. The
FEniCS form compiler (FFC) \cite{logg2012ffc} compiles code written in
UFL to Unified Form-assembly Code (UFC) \cite{alnaes2012ufc} which are
optimized C++ code. The Python interface also makes use of the Instant
module which allows for just-in-time (JIT) compilation of C++
code. The compiled code is also stored in a cache so that compilation
of a form only happens once. Also, the relatively new UFL Analyser and
Compiler System (UFLACS) allows for fast compilation of complex forms
such as variational formulations that include the Holzapfel Ogden
material model \eqref{eq:holzapel_full}.

@@

An even more clever strategy uses a technique from bifurcation and
chaos theory and is known as numerical continuation
\cite{allgower2003introduction}.  Suppose we want to
solve the non-linear problem $F(\uvec, \lambda)=0$ with state variable
$\uvec$ and parameter $\lambda$. For instance $\uvec$ could be the
displacement and $\lambda$ could be the endocardial pressure.
The idea behind numerical continuation is that given a solution pair
$(\uvec_0, \lambda_0)$ there exist (under conditions stated by the
implicit function theorem) a solution curve $\uvec(\lambda)$ such that
$F(\uvec(\lambda), \lambda)=0$ and $\uvec(\lambda_0) = \uvec_0$.
To explicitly find such a curve is not always easy but a simple
approximation can be found by extrapolation: Given two pairs
$(\uvec_0, \lambda_0)$ and $(\uvec_1, \lambda_1)$, and a new target
parameter $\lambda_2$, a possible solution is 
\begin{align}
  \uvec_2 =  (1-\delta)\uvec_0 + \delta \uvec_1 && \delta = \frac{\lambda_2 - \lambda_0}{\lambda_1 - \lambda_0}.
\end{align}
% Let $w_0$ denote the initial
% state variable associated with endocardial pressure $p_0$. Increase
% the pressure $p_1 = p_0 + \Delta p_0$, and solve to obtain $w_1$. Next
% we would like to solve for $p_2 = p_1 + \Delta p_1$ where $\Delta
% p_1$ might be the adapted step size. In stead of using $\tilde{w_2}=w_1$ as
% initial guess for the newton solver, as we typically would do in the incremental load
% technique, we observe that if $\delta = \frac{p_2 - p_0}{p_1 - p_0}$,
% and hence $p_2 = (1-\delta)p_0 + \delta p_1$, then a better choice
% of intial guess would be $\tilde{w_2} = (1-\delta)w_0 + \delta w_1$.
Choosing $\uvec_2$ as initial guess for the non-linear solver has been
successfully performed by others in non-linear cardiac mechanics
problems \cite{pezzuto2013mechanics}, and this approach is also used
in this thesis.

>>>>

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants \cite{wang1970new} states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is, the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T = \mathbf{a}_0\otimes\mathbf{a}_0 \},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
Some of the invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

@@

The myocardium contains small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts, this
perfused blood is squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. A material that change its
volume in response to applied loads is referred to as compressible. When the 
volume is preserved we say that the material is incompressible.
Since 2-4\% is very little, a common assumption in cardiac mechanical modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompressible. The reason for this choice is
purely numerical.

@@

\subsection{Force-balance equation}
We will now collect all the terms that are involved in the force balance for
the cardiac mechanics problem. Considering the myocardium as an incompressible,
hyperelastic material we obtain the following strong form in the
Lagrangian formulation 
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derive the weak
variational form of this equation.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

In this thesis we have used a transversely isotropic version of
\eqref{eq:holzapel_full} which is obtained by setting $a_{fs} =
b_{fs}= a_s = b_s = 0$, i.e 
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces (after a
series expansion of the exponential and a limiting argument) to
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    J\Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right),
  \end{split}
\end{align}
where $\mathbf{B} = \F \F^{T}$ is the left Cauchy-Green tensor,
$\mathbf{f} = \F \ef$ and $\mathbf{s} = \F \eS$.

@@

The active part $\F_a$, is an inelastic process driven by the
biochemistry and can be seen as the actual distortion of the
microstructure. The elastic part $\F_e$ is responsible for preserving
compatibility of the tissue and stores all the energy in the
deformations. As a consequence, the strain energy function is a
function of the elastic deformation gradient only. The
decoupling can be illustrated by considering two sarcomeres connected in
series as shown in Figure \ref{fig:actstrain}.

@@

The FEniCS project is an open-source computing platform for solving
partial differential equations (PDEs) using the finite element method (FEM).
Solving PDEs using FEM involves many implementation details that can
be tedious to implement yourselves. The idea behind FEniCS is to
automate code generation so that the user can spend more time on doing
research and less time on implementation of assembly matrices. At the
core of FEniCS is DOLFIN \cite{logg2012dolfin}, which is a C++/Python
library, and works as the main interface in FEniCS. In this thesis
only the Python interface has been used, in which C++ code is
automatically generated using SWIG. This allows for simplicity through
the Python scripting language and the speed of the C++ language.
The domain specific language used to represent weak formulations is
called the Unified Form Language (UFL) \cite{alnaes2014unified}, and
allows for e.g automatic differentiation of forms and expressions. The
FEniCS form compiler (FFC) \cite{logg2012ffc} compiles code written in
UFL to Unified Form-assembly Code (UFC) \cite{alnaes2012ufc} which are
optimized C++ code. The Python interface also makes use of the Instant
module which allows for just-in-time (JIT) compilation of C++
code. The compiled code is also stored in a cache so that compilation
of a form only happens once. Also, the relatively new UFL Analyser and
Compiler System (UFLACS) allows for fast compilation of complex forms
such as variational formulations that include the Holzapfel Ogden
material model \eqref{eq:holzapel_full}.

@@

An even more clever strategy uses a technique from bifurcation and
chaos theory and is known as numerical continuation
\cite{allgower2003introduction}.  Suppose we want to
solve the non-linear problem $F(\uvec, \lambda)=0$ with state variable
$\uvec$ and parameter $\lambda$. For instance $\uvec$ could be the
displacement and $\lambda$ could be the endocardial pressure.
The idea behind numerical continuation is that given a solution pair
$(\uvec_0, \lambda_0)$ there exist (under conditions stated by the
implicit function theorem) a solution curve $\uvec(\lambda)$ such that
$F(\uvec(\lambda), \lambda)=0$ and $\uvec(\lambda_0) = \uvec_0$.
To explicitly find such a curve is not always easy but a simple
approximation can be found by linear extrapolation: Given two pairs
$(\uvec_0, \lambda_0)$ and $(\uvec_1, \lambda_1)$, and a new target
parameter $\lambda_2$, a possible solution is 
\begin{align}
  \uvec_2 =  (1-\delta)\uvec_0 + \delta \uvec_1 && \delta = \frac{\lambda_2 - \lambda_0}{\lambda_1 - \lambda_0}.
\end{align}
% Let $w_0$ denote the initial
% state variable associated with endocardial pressure $p_0$. Increase
% the pressure $p_1 = p_0 + \Delta p_0$, and solve to obtain $w_1$. Next
% we would like to solve for $p_2 = p_1 + \Delta p_1$ where $\Delta
% p_1$ might be the adapted step size. In stead of using $\tilde{w_2}=w_1$ as
% initial guess for the newton solver, as we typically would do in the incremental load
% technique, we observe that if $\delta = \frac{p_2 - p_0}{p_1 - p_0}$,
% and hence $p_2 = (1-\delta)p_0 + \delta p_1$, then a better choice
% of intial guess would be $\tilde{w_2} = (1-\delta)w_0 + \delta w_1$.
Choosing $\uvec_2$ as initial guess for the non-linear solver has been
successfully performed by others in non-linear cardiac mechanics
problems \cite{pezzuto2013mechanics}, and this approach is also used
in this thesis.

### 2017-12-20 chapters/introduction/patient.tex
Update figures. Change size and convert to eps
Fix some final typos

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used together with the medical images to
  assimilate additional data in order to extract clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh_plane.pdf}
    \caption{\label{fig:strain_mesh_plane}}
  \end{subfigure}
  \begin{subfigure}[t]{0.36\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/cut.png}
    \caption{\label{fig:cut_intro}}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/mesh.png}
    \caption{\label{fig:mesh_intro}}
  \end{subfigure}
\caption{A least square fitting plane is fitted to the points
  belonging to the basal boundary in the strain mesh (Figure
  \ref{fig:strain_mesh_plane}). This plane is used to cut the endo- and
  epicardial surfaces (Figure \ref{fig:cut_intro}) and Gmsh is used to mesh
  these surfaces together (Figure \ref{fig:mesh_intro}). }
\label{fig:mesh_generation_intro}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??
A general problem in biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of an inflated version of the balloon. Having
this picture in mind, it is intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. Figure
\ref{fig:unloading_lv} illustrate the backward displacement method applied
to an LV geometry. We briefly explain the steps in this methods.

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of finding the unloaded, zero pressure geometry.
In \cite{nikou2016effects}, Nikou et al proposed a method for estimating
material parameters and the unloaded geometry by iterating between
the two methods, and terminating if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters. In the last paper in this thesis, we analyze this method
and show that it actually converges to the correct solution and is
stable to noise in the measurements.

@@

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{strain}}$}
Another observation that is encountered in this thesis are strain
measurements, or more precisely average regional strain in the
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vector field in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Green-Lagrange strain tensor $\mathbf{E}$, or
the material displacement gradient tensor $\F - \I$.

@@

Another reason for using regularization is to stabilize the solver for
the underlying PDE that we solve. Parameters with sharp spikes
can make Newtons iterations in-feasible, and restricting the
parameter space to exclude such solutions might help stabilize the
Newton solver.

@@

Demonstrating identifiability is 
often difficult, especially when your data is corrupted with noise,
and when the complexity of the model increases. Balancing the
complexity of the model in terms of \emph{model fidelity}, i.e whether
your 
model is rich enough to representing the data, and the identifiability
is often key in order to have robust model. A too simple model will
often fail to represent your data, while a too complex model might
give completely different answers with just a slight perturbation of
the data, i.e with noise added to the measurement. The same problem is
encountered in other scientific disciplines. For example, in
statistical learning theory, \emph{overfitting} is related to
overparameterization of the model, in which you can perfectly
represent your data, but a small perturbation will cause a significant
error.

>>>>

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used together with the medical images to
  assimilate additional data in order to extract clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh_plane}
    \caption{\label{fig:strain_mesh_plane}}
  \end{subfigure}
  \begin{subfigure}[t]{0.36\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/cut.png}
    \caption{\label{fig:cut_intro}}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/mesh.png}
    \caption{\label{fig:mesh_intro}}
  \end{subfigure}
\caption{A least square fitting plane is fitted to the points
  belonging to the basal boundary in the strain mesh (Figure
  \ref{fig:strain_mesh_plane}). This plane is used to cut the endo- and
  epicardial surfaces (Figure \ref{fig:cut_intro}) and Gmsh is used to mesh
  these surfaces together (Figure \ref{fig:mesh_intro}). }
\label{fig:mesh_generation_intro}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??
A general problem in biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning that the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of an inflated version of the balloon. Having
this picture in mind, it is intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. Figure
\ref{fig:unloading_lv} illustrate the backward displacement method applied
to an LV geometry. We briefly explain the steps in this methods.

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of finding the unloaded, zero pressure geometry.
In \cite{nikou2016effects}, Nikou et al proposed a method for estimating
material parameters and the unloaded geometry by iterating between
the two methods, and terminating if the difference in unloaded cavity
volume was less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters. In the last paper in this thesis, we analyze this method
and show that it actually converges to the correct solution and is
stable to noise in the measurements.

@@

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{strain}}$}
Another observation that is encountered in this thesis is strain, or
more precisely average regional strain in the 
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vector field in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V.
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Green-Lagrange strain tensor $\mathbf{E}$, or
the material displacement gradient tensor $\F - \I$.

@@

Another reason for using regularization is to stabilize the solver for
the underlying PDE that we solve. Parameters with sharp spikes
can make Newtons iterations in-feasible, and restricting the
parameter space to exclude such solutions might help to stabilize the
Newton solver.

@@

Demonstrating identifiability is 
often difficult, especially when your data is corrupted with noise,
and when the complexity of the model increases. Balancing the
complexity of the model in terms of \emph{model fidelity}, i.e whether
your 
model is rich enough to represent the data, and the identifiability
is often key in order to have robust model. A too simple model will
often fail to represent your data, while a too complex model might
give completely different answers with just a slight perturbation of
the data, i.e with noise added to the measurement. The same problem is
encountered in other scientific disciplines. For example, in
statistical learning theory, \emph{overfitting} is related to
overparameterization of the model, in which you can perfectly
represent your data, but a small perturbation will cause a significant
error.
